---
title: "lme4qtl: an efficient and flexible QTL mapper"
subtitle: "The ESHG 2016 conference, Barcelona, Spain"
author: Andrey Ziyatdinov
date: May, 2016
output: ioslides_presentation
---


```{r setup, echo = FALSE, message = FALSE, cache = FALSE}
# set global chunk options
opts_chunk$set(fig.path = 'figure/', cache.path = 'cache/', echo = FALSE, cache = TRUE, 
  tidy = FALSE, 
  fig.width = 9, fig.height = 4.5, dev = 'png',
  warning = FALSE, message = FALSE)
# upload images automatically? (identity/imgur_upload)
opts_knit$set(upload.fun = identity)
```

```{r inc, cache = FALSE}
library(devtools)
load_all("~/git/hemostat/lme4qtl")
```

```{r dat}
source("R/01-load-data.R")
```

```{r include_cite, cache = FALSE}
#library(knitcitations)
```

```{r knitcitations, cache = FALSE}
#cite_options(linked = TRUE)
#cite_options(tooltip = TRUE)

#bib <- read.bibtex("mendeley.bib")
```

## lme4qtl is for you if

* You use mixed models for QTL mapping 
    * Basic models (efficiency)
    * Advanced models (flexibility)
* You code in R
    * If not, take a look at the formula interface
* You are a fan of the [lme4](https://github.com/lme4/lme4) R package
    * If not, [dmbates](https://github.com/dmbates) is watching you


<div style="float: right; font-size: 50%;">
![](https://avatars2.githubusercontent.com/u/371258?v=3&s=150)
<br/>
Source: [github.com/dmbates](https://github.com/dmbates)
</div>

# Implementation

## 2-column Layout

<div class="columns-2">
  - Bullet 1
  - Bullet 2
  - Bullet 3

  image.png
</div>

# Results | QTL mapping examples using lme4qtl

## Models

* Polygenic (quantitative trait)
* Polygenic (binary trait)
* Polygenic (counts)

<br/>

* Association (quantitative trait)
* Linkage (quantitative trait)

<br/>

* GxE (sex-especificity)
* GxE (ageing)

## Description of the GAIT data {.smaller}

|  R objects |  Description  |
|:-----------|:--------------|
| phen/phen2 | A data frame with phenotypes and IDs (GAIT1/GAIT2) |
| dkin/dkin2 | The double kinship matrix with IDs in row/columns names (GAIT1/GAIT2)|

## Description of the GAIT data (variables) {.smaller}

|  Variable  | Type | Project | Description  |
|:-----------|:-----|:--------|:-------------|
| ID  | character | |The individual's ID (must be unique)|
| HHID | character | |The individual's house-hold ID |
| SEXf | factor | |The gender (Male/Female) |
| AGE/AGEc/AGEsc |numeric | |The age (raw/centered/scaled) |
| AGEc2/AGEsc2 | numeric | |The age squared (centered/scaled) |
| APTT | numeric | GAIT1 | Activated Partial Thromboplastin Time (units, sec) |
| Throm | factor | GAIT2 | The thrombosis disease status (control/affected) |

## Polygenic model for APTT

```{r aptt, echo = T}
m <- relmatLmer(APTT ~ AGEsc + (1|ID), phen, relmat = list(ID = dkin))
m
```

## Residuals

```{r aptt_res, echo = T}
r <- residuals(m)
```

<div class="columns-2">

```{r aptt_res_1, fig.width = 4, fig.height = 4, echo = T}
qqnorm(r); qqline(r)
```

```{r aptt_res_2, fig.width = 4, fig.height = 4, echo = T}
hist(r, breaks = 30)
```

</div>

## Polygenic model for Throm

```{r m1, echo = TRUE}
m1 <- relmatGlmer(Throm ~ (1|ID), phen2, relmat = list(ID = dkin2), 
  family = binomial)
m1
```


## References

```{r, results = "asis"}
#bibliography()
```

## Thank you
